---
title: World production and grain composition of major cultivated species
author: Deependra Dhakal
date: '2019-02-03'
slug: world-production-and-grain-composition-of-major-cultivated-species
categories:
  - agriculture
  - Plant Breeding
tags:
  - agriculture
  - plant breeding
header:
  caption: ''
  image: ''
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(tidy = TRUE, cache = FALSE, echo = FALSE, 
                      tidy.opts = list(width.cutoff=50), 
                      eval = TRUE, fig.align = "center", fig.asp = 0.6, 
                      message = FALSE, warning = FALSE, cache = TRUE)
options(knitr.kable.NA = "", digits = 3, knitr.table.format = "html")
library(tidyverse)
```

# Context

Out of crops raised for their seed/grains (listed under 35 species, by FAO; FAOSTAT, 2014), only 22 species are produced in substantial amounts. Species of graminae and leguminosae families alone account for about 85 percent of the total grain production. As presented here in Table \@ref(tab:world-production).

```{r world-production-major-species}
prod_comp <- read_csv(here::here("content", "blog", "data", "world_production_major_species.csv"), skip = 2)
```


## Production

```{r world-production}
prod_comp %>% 
  select(Crop, `Crop species`, `World production^[Average of 2011 to 2014, FAOSTAT (2016)] (1000 t)`) %>% 
  knitr::kable(caption = "Global production of major cultivated crops", format = "html") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12) %>%
  kableExtra::column_spec(1, bold = TRUE) %>% 
  kableExtra::scroll_box(width = "100%", 
                         box_css = "border: 2px solid #22d0dd; padding: 10px; ")
```

## Grain composition

```{r grain-comp}
prod_comp %>% 
  select(-`World production^[Average of 2011 to 2014, FAOSTAT (2016)] (1000 t)`) %>% 
  knitr::kable(caption = "Global production of major cultivated crops", format = "html") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12) %>%
  kableExtra::column_spec(1, bold = TRUE) %>% 
  kableExtra::scroll_box(width = "100%", 
                         box_css = "border: 2px solid #22d0dd; padding: 10px; ")
```


# References

Page 3 and 4, Seed Biology and Yield of Grain Crops, 2nd Edition

Seed composition data from Bewley et.al. (2013), Sinclair and de Wit (1975), Langer and Hill (1991), Hulse et.al. (1980), and Alberta Agriculture and Forestry (2015)
