---
title: Dealing with factors
author: Deependra Dhakal
date: '2017-05-18'
slug: []
categories:
  - R
  - factor
  - blog
tags:
  - R
  - factors
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE, 
                      tidy.opts = list(width.cutoff=50), 
                      eval = TRUE, fig.align = "center", fig.asp = 0.6, 
                      message = FALSE, warning = FALSE)
options(knitr.kable.NA = "", digits = 3)
library(tidyverse)
```

## A `factor` is an headache

I have a dataset, cleaning which has been a pain lately. I'm going to use 20 observations of the imported dataset in this post to demonstrate how pathetically have I been advancing with it. 

```{r import, echo=FALSE, results='asis'}
head_boot <- read_csv(here::here("content", "blog", "data", "factor_data.csv"), na = "") %>% 
  mutate_if(is.character, factor)
# str(head_boot)
knitr::kable(head(head_boot), format = "html", align = "c") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), 
                            font_size = 10, position = "center") %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::column_spec(1, bold = TRUE)
```

Providing it a context, the columns represent multiple observations of same variable at different dates, as apparent from the column names.

One can observe from the `str(head_boot)` that the dataframe has only 20 observations. Had a larger dataframe been subsetted, the factors with their usual nuisance carry over the all the levels, even currently unused ones. To demonstrate this, below I present how many factor levels the original dataframe had and what changes after a `subset` (`slice`, `filter` have same feature) operation.


```{r unused-factor-levels, echo=TRUE}
# original dataframe
head_boot %>% 
  select_if(is.factor) %>% 
  map_int(function(x)length(levels(x)))

# subsetted to obtain a dataframe where one factors has either "s" or "0" levels only
subset.data.frame(head_boot, 
                  subset = head_boot$jan_23_2017 %in% c("s", "0"), 
                  select = sapply(head_boot, is.factor)) %>% 
  map_int(function(x)length(levels(x)))

```


All factor levels are retained ! 

A prophylactic measure is avoiding at all importing of character data columns as a factor. This could be done using setup option: `options(stringsAsFactors = FALSE)` ahead of import. Still, after the import many options exist. Some of the best alternatives out are listed below:

1. Use `subsetted_df$factor_col[, drop=TRUE]`, where a vector of factor values is indexed with drop argument.

2. Convert factor columns of subsetted dataframe first to character and then back to factors class.

3. Use `forcats::fct_drop(f)` for more reliable NA handling

4. Use `droplevels()` or `droplevels.data.frame()`


To give a realistic feel to data, factor levels may be recoded or relabelled with more informative labels.

```{r relabel-factors, echo=TRUE}
suit_labels <- factor(c("before_boot_n_boot", "heading", "anthesis"), 
                      levels = c("before_boot_n_boot", "heading", "anthesis"), 
                      ordered = T)

# what are all the levels
head_boot_lvl <- head_boot %>% 
  select_if(is.factor) %>% 
  map(levels) %>% 
  unlist() %>% 
  unique()

# create all combinations of factor levels of the dataset
# purrr::cross_df preserves the order, tidyr::crossing doesn't
possible_hb_lvl <- cross_df(list(post  = factor(c("m", "", "p"), ordered = TRUE), 
      pre = factor(c("b", "f", "a", "s", "1", "2", 
                     "3", "4", "10", "11", "12"), ordered = TRUE))) %>% 
  select(rev(everything())) %>% 
  # arrange(match(post, c("m", "", "p"))) %>% # don't need this ordering 
  unite(col = "pre_post", sep = "") %>% 
  add_row(pre_post = "0", .before = 1) %>% 
  mutate(pre_post = factor(pre_post, levels = pre_post, ordered = TRUE))

# mutate original dataframe to represent ordinal factors
head_boot <- head_boot %>% 
  mutate_if(is.factor, function(x)factor(x, levels = possible_hb_lvl$pre_post, labels = possible_hb_lvl$pre_post, ordered = TRUE))

# function to recode factor levels (better to use `fct_recode()`)
fct_minimize <- function(x)case_when(x < "1m" ~ suit_labels[1], 
                                     x >= "1m" & x < "10m" ~ suit_labels[2], 
                                     x >= "10m" ~ suit_labels[3], 
                                     TRUE ~ NA)

# mutate original dataframe with function to recode factor levels
head_boot <- head_boot %>% 
  mutate_if(is.factor, fct_minimize)

str(head_boot)
```

A more telling dataframe looks like the one below.

```{r results='asis'}
knitr::kable(head(head_boot), format = "html", align = "c") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), 
                            font_size = 10, position = "center") %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::column_spec(1, bold = TRUE)
```


## Visualizing data

Bar plotting of the ordinal data might reveal interesting insights, so let's prepare ggplot graphs.

```{r ggbar-viz, fig.cap="Growth stages at different dates"}
walk2(.x = head_boot %>% select(jan_23_2017:feb_02_2017) %>% colnames(),
      .y = as.character(strptime(str_subset(colnames(head_boot), "^\\w{3}_\\d{2}"), 
                                 format = "%b_%d_%Y")),
      .f = ~print(ggplot(aes(x = get(.x)), data = head_boot) +
                    geom_bar(position = "dodge", stat = "count", width = .5) +
                    xlab(.y) +
                    theme(axis.text.x = element_text(angle = 40, size = 9, vjust = 0.6))))
```

With dplyr, It's seemless to obtain a tabular summary of what has been visualized above.

```{r hb-summ-tab, results='asis'}
# credit: https://stackoverflow.com/a/46340237/6725057
head_boot %>% 
  select_if(is.factor) %>% 
  tidyr::gather(date, stage) %>% 
  dplyr::group_by(date, stage) %>% 
  dplyr::count() %>% 
  dplyr::ungroup() %>% 
  tidyr::spread(date, n) %>% 
  knitr::kable(format = "html", align = "c") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), 
                            font_size = 10, position = "center") %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::column_spec(1, bold = TRUE)
```

