---
title: Variance component based parameter estimation of incomplete block designs
author: Deependra Dhakal
date: '2018-12-12'
slug: variance-component-based-parameter-estimation
categories: 
  - R
tags: 
  - R
header:
  caption: ''
  image: ''
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(tidy = TRUE, cache = FALSE, echo = FALSE, 
                      tidy.opts = list(width.cutoff=50), 
                      eval = TRUE, fig.align = "center", fig.asp = 0.6, 
                      message = FALSE, warning = FALSE, cache = TRUE)
options(knitr.kable.NA = "", digits = 3, knitr.table.format = "html")
library(tidyverse)
library(agricolae)
set.seed(453)
```


## Introduction

Variance component models are also suited for analysis of incomplete block designs, besides complete block designs. This post aims to demonstrate exactly that. Using a dataset generated from alpha lattice design, I show how the design can be properly modeled and fit using OLS regression having various fixed model components. This system of model fitting is analogous to classical ANOVA based technique of estimating parameters.

## The dataset

The data is same as introduced earlier in "Design and analysis of balanced randomized complete block (RCBD) experiment" [post](https://rookie.rbind.io/post/design-and-analysis-of-balanced/). It comprises of plant height trait for the maize recorded in multiple replication units, only difference from the earlier post being that the arrangement of observation plots into blocks which are themselves nested within replication will be accounted for. This is the objective of this post: to account for effects of incomplete blocks which was incorporated in the design.

An overview of how the data looks like is shown in Table \@ref(tab:ihyb-height).

```{r ihyb-height}
ihyb_multiple <- read_csv(here::here("content", "blog", "data", "ihyb_2018.csv")) %>% 
  mutate_at(c("Rep", "Block", "Plot", "Entry", "col", "row"), as.factor) %>% 
  mutate(`Plant height` = rowMeans(select(., contains("Plant height")), na.rm = TRUE)) %>% 
  select(which(!str_detect(colnames(.), "height\\d")), -tillering, -moisture1, -moisture2, -`Ear count`, -row, -col) %>% 
  rename(Plant_height = `Plant height`)

ihyb_multiple %>% 
  head(n = 8) %>% 
  knitr::kable(caption = "Intermediate maturing hybrids with 50 entries each in 3 replicated blocks") %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", font_size = 10, full_width = TRUE) %>% 
  kableExtra::row_spec(0, bold = TRUE)
```

## Model formulation

```{r model-formulation, echo=TRUE}
model_pht <- lm(formula(paste("Plant_height", "~ Rep + Entry + Block%in%Rep")), 
            data = ihyb_multiple)
ANOVA_pht <- anova(model_pht) # model anova
```

```{r anova-print}
# print anova
ANOVA_pht %>% 
  knitr::kable(caption = "ANOVA of fixed effects factors with block nested within replication") %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", font_size = 10, full_width = TRUE) %>% 
  kableExtra::row_spec(0, bold = TRUE)
```


```{r model-fitting}
# parameter initialization
obs <- nrow(ihyb_multiple) # number of observations
ntr <- nlevels(ihyb_multiple$Entry) # number of entries
nrep <- nlevels(ihyb_multiple$Rep) # number of replications
nblock <- nlevels(ihyb_multiple$Block) # number of blocks
trt <- ihyb_multiple$Entry # entry vector
replication <- ihyb_multiple$Rep # replication vector
block <- ihyb_multiple$Block # block vector
y <- ihyb_multiple$Plant_height # response vector
sblock <- obs/nblock # size of block
glt <- ntr-1
s <- ntr/sblock # number of blocks per replication
b <- s * nrep

mean_pht <- mean(ihyb_multiple$Plant_height, na.rm = TRUE) # mean of response
df_resid_pht <- df.residual(model_pht) # residual degrees of freedom
SCt <- anova(model_pht)[2, 2] # sum of squares of treatment 
Ee <- deviance(model_pht)/df_resid_pht # mean sum of square of error
Eb <- anova(model_pht)[3, 3] # mean sum of square of rep:block 
X <- rep(0, obs * ntr) # initialize matrix of 0 values for matrix of entry/treatment
dim(X) <- c(obs, ntr) # change dimension of matrix to generate model matrix for entry/treatment
for (i in 1:obs) { # design of model matrix for entry/treatment
  tr <- trt[i]
  X[i, tr] <- 1
}
R <- rep(0, obs * nrep) # initialize matrix of 0 values for matrix of replication
dim(R) <- c(obs, nrep) # change dimension of matrix to generate model matrix for replication
for (i in 1:obs) { # design of model matrix for replication
  rp <- replication[i]
  R[i, rp] <- 1
}
Z <- rep(0, obs * nblock) # initialize matrix of 0 values for matrix of blocks
dim(Z) <- c(obs, nblock) # change dimension of matrix to generate model matrix of blocks
for (i in 1:obs) { # design of model matrix for blocks
  rb <- block[i]
  Z[i, rb] <- 1
}
N <- t(X) %*% Z # matrix multiplication of treatment structure-matrix with block structure-matrix
In <- diag(1, obs) # diagonal matrix of observation structure (150 x 150)
c0 <- t(Z) %*% (In - (1/nrep) * X %*% t(X)) %*% y # column vector of 30 x 1 dim
Js <- diag(s) # blocks per replication diagonal matrix
Ir <- diag(nrep) # replication diagonal matrix
Jr <- matrix(1, nrep, nrep) # unity matrix of 3 x 3 dimension
Js <- matrix(1, s, s) # unity matrix of 10 x 10 dim
Ib <- diag(b) # block diagonal matrix
Iv <- diag(ntr) # treatment diagonal matrix
q <- sblock - floor(sblock/s) * s 

if (q <= s/2) {g <- floor(sblock/s)}
if (q > s/2) {g <- floor(sblock/s) + 1}

phi <- nrep * (Eb - Ee)/((nrep - 1) * Ee)
lambda <- 1/(nrep * sblock * (1/phi + 1) - sblock)
W <- t(N) %*% N - sblock * Ib - g * kronecker((Jr - Ir), Js)
# (t(N) %*% N) %>% dim # matrix of dimension 30 x 30
# (sblock * Ib) %>% dim # matrix of dimension 30 x 30
# (kronecker((Jr-Ir), Js)) %>% dim # matrix of dimension 30 x 30

inversa <- MASS::ginv(Ib - lambda * W)
tauIntra <- t(X) %*% y/nrep - lambda * N %*% inversa %*% c0
vartau <- (Ee/nrep) * (Iv + lambda * N %*% inversa %*% t(N))
dvar <- sqrt(diag(vartau))
```


```{r vc-model-results}
# set parameters to omit na values and get common statistics
ntr0 <- ncol(vartau)
if (ntr0 < ntr) {
  ntr <- ntr0
  nrep <- na.omit(nrep) # na omitted levels of replication
}

# difference ???
vardif <- matrix(0, ntr, ntr)
for (i in 1:(ntr - 1)) {
  for (j in (i + 1):ntr) {
    vardif[i, j] <- vartau[i, i] + vartau[j, j] - 2 * vartau[i, j]
    vardif[j, i] <- vardif[i, j]
  }
}

comb <- utils::combn(ntr, 2)
nn <- ncol(comb)
dif <- rep(0, nn)
stdt <- rep(0, nn) # initialize standard error
pvalue <- rep(0, nn)

# compute p-value and standard error for significance of difference between two treatment means
for (k in 1:nn) {
  i <- comb[1, k]
  j <- comb[2, k]
  dif[k] <- tauIntra[i] - tauIntra[j]
  stdt[k] <- sqrt(vartau[i, i] + vartau[j, j] - 2 * vartau[i,j]) # standard error
  tc <- abs(dif[k])/stdt[k]
  pvalue[k] <- 2 * round(1 - pt(tc, df_resid_pht), 4) # lsd test
  # pvalue[k] <- round(1 - ptukey(tc, ntr, df_resid_pht),4) # tukey test
}

# pvalue values' vector
tr.i <- comb[1, ]
tr.j <- comb[2, ]
Q <- matrix(1, ncol = ntr, nrow = ntr)
p <- pvalue

# probability values' matrix
k <- 0
for (i in 1:(ntr - 1)) {
  for (j in (i + 1):ntr) {
    k <- k + 1
    Q[i, j] <- p[k]
    Q[j, i] <- p[k]
  }
}

```


```{r overall-mean-and-model-cv, echo=TRUE}
mean_pht # overall mean

Fstat <- data.frame("Fit Statistics" = c(AIC = AIC(model_pht), BIC = BIC(model_pht)))
Fstat # fit statistics

E <- (ntr - 1) * (nrep - 1)/((ntr - 1) * (nrep - 1) + nrep * (s - 1)) 
# ntr = number of treatments, 
# nrep = number of replications
# s = number of blocks per replication 
E # efficiency factor

CV <- sqrt(Ee) * 100/mean_pht # where, Ee = deviance(model_pht)/df_resid_pht 
CV # model CV 
```


```{r model-means-and-group}
# mean and groups of treatments
groups <- agricolae::orderPvalue(treatment = 1:ntr, means = tauIntra, alpha = 0.05, pvalue = Q, console = FALSE)
groups <- groups %>% rownames_to_column("treatment")

# print mean and group
groups %>% 
  head(n = 15) %>% 
  knitr::kable(caption = "Treatment means and groups for the response variable") %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", font_size = 10, full_width = TRUE) %>% 
  kableExtra::row_spec(0, bold = TRUE)
```


Alternatively, analysis and results from analysis could be obtained via agricolae function `PBIB.test()`. 

```{r verify-agricolae, echo=TRUE}
# vc_model_alpha <- agricolae::PBIB.test(block = block, trt = trt, replication = replication, 
#                      k = sblock, y = y, method = "VC", console = TRUE)
# vc_model_alpha$groups
```

