---
title: "Varietal database of lentil in Nepal of Nepal"
author: "Deependra Dhakal"
date: '2018-10-26'
header:
  caption: ''
  image: ''
slug: varietal-database-of-lentil-in-nepal
tags:
- agriculture
- research-review
categories:
- agriculture
- Plant Breeding
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(tidy = TRUE, cache = TRUE, 
                      echo = FALSE, 
                      tidy.opts = list(width.cutoff=50), 
                      eval = TRUE, 
                      fig.show = "hold", fig.align = "center", fig.width = 6,
                      fig.asp = 0.8, message = FALSE, warning = FALSE)
options(knitr.kable.NA = "", digits = 3)
```

This one is my effort to compose an updated database on the current situation of lentil germplasm in Nepal. I've managed to list out the varieties that have been made available so far (either through release or registration process). Although, a lot of other popular genotypes are trending in cultivation as of now. I consider NARC's varietal catalog to be the most authentic, so have borrowed most of the information from the published documents.

Several inconsistencies exists even among higly praised documents, mainly for the attributes that are most affected by environmental variation; these include, but not limited to, yield, days to flowering and days to maturity traits. Besides that, publishers should also have assisted the records with the mention of cropping system in which the varieties were raised. It's implication cannot be underemphasized, and real case scenario could justify that. For example, a crop certainly yield much in a sole cropped situation in contrast to when it is grown as mixed crop. A reasonable comparison between these two systems of cultivation cannot be made precisely. So, it's best to leave with an explaination about how crops were grown in these types of situation.  Wherever possible these values should be further contextualized (for example by, noting the actual location it were grown in, the season it was sown, etc.). 

As for this post, the database will to be updated as the information comes along to reflect the latest details. Following is a list of released and registered varieties of lentil:

```{r varieties-list}
lentil_vars <- read_csv("https://raw.githubusercontent.com/DeependraD/publicdata/master/lentil_varieties.csv")

lentil_vars %>%
  knitr::kable(caption = "Released and registered varieties of lentil in Nepal") %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", font_size = 10) %>%
  kableExtra::column_spec(1, bold = TRUE)
```

It it very unfortunate that we have very contrasting information published (without explaination) in different versions of national databases. For instance, in Table \@ref(tab:varieties-list), `Production1 (ton_per_ha)` column was borrowed from "Krishi Diary, 2075" and the other column for the same yield trait (`Production2 (ton_per_ha)`) from elsewhere. I defer the exact reference and citation for a later date.
