---
title: Diallel cross analysis
author: Deependra Dhakal
date: '2018-09-15'
slug: diallel-cross-analysis
math: true
bibliography: ["./bib/2018-09-15-diallel-cross-analysis.bib"]
link-citations: true
categories:
  - Plant Breeding
tags:
  - agriculture
  - research-review
header:
  caption: ''
  image: ''
---


```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(tidy = TRUE, cache = TRUE, echo = FALSE, 
                      tidy.opts = list(width.cutoff=50), 
                      eval = TRUE, fig.align = "center", fig.asp = 0.6, 
                      message = FALSE, warning = FALSE)
options(knitr.kable.NA = "", digits = 3, knitr.table.format = "html")
library(tidyverse)
require(plantbreeding)
```

## Mating designs: a brief review

Mating designs allow for partitioning of effects of phenotype, as either due to genotype, environment or dominance effects.

Using one or more analysis scheme information such as identification of heterotic groups, estimation of general and specific combining abilities and testing of environmental interactions can be extracted from mating designs. Progenies resulting from a well designed mating are used for the study of genetic traits.

There are four types of popular diallel crosses:

1. Full diallels: All the possible combinations of crosses among parents, including reciprocals and self-fertilization of the parents are made. For a sample of $n$ parents, the full-diallel requires $n \times n$ progenies, a number that quickly becomes unmanageable as more parents are sampled (Table \@ref(tab:full-diallel)).

```{r full-diallel}
# Full diallel mating scheme

ten_ind <- gl(10, k = 1, ordered = T, labels = paste("P", 1:10, sep = ""))
mating_mat <- cross_df(.l = list(M = ten_ind, F = ten_ind)) %>% 
  unite(col = "cross", sep = " x ", remove = F) %>% 
  spread(key = F, value = cross) %>% 
  dplyr::rename(Parents = M)

mating_mat %>%
  knitr::kable(caption = "Full diallel mating scheme using 10 parents") %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", font_size = 10) %>% 
  kableExtra::column_spec(1, bold = TRUE) %>% 
  kableExtra::column_spec(column = 1:11, width = "7em")
```


2. Half diallels: Each parent is mated with every other parent, excluding selfs and reciprocals. This requires making $\frac{n(n-1)}{2}$ crosses for n parents (Table \@ref(tab:half-diallel)).

```{r half-diallel}
# Half diallel mating scheme

ten_ind <- gl(10, k = 1, ordered = T, labels = paste("P", 1:10, sep = ""))
mating_mat <- cross_df(.l = list(M = ten_ind, F = ten_ind)) %>% 
  unite(col = "cross", sep = " x ", remove = F) %>% 
  spread(key = F, value = cross) %>% 
  dplyr::rename(Parents = M)

mating_mat[,-1][upper.tri(mating_mat[,-1], diag = TRUE)] = NA
  
mating_mat %>%
  knitr::kable(caption = "Half diallel mating scheme using 10 parents") %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", font_size = 10) %>% 
  kableExtra::column_spec(1, bold = TRUE) %>% 
  kableExtra::column_spec(column = 1:11, width = "7em")
```


3. Partial diallel: Not all the crosses are made. There are no reciprocals or selfs. The goal is to reduce the breeding workload for a given sample of parents by making less than $\frac{n(n-1)}{2}$ crosses for n parents (Table \@ref(tab:partial-diallel)).

```{r partial-diallel}
# Partial diallel mating scheme

ten_ind <- gl(10, k = 1, ordered = T, labels = paste("P", 1:10, sep = ""))
mating_mat <- cross_df(.l = list(M = ten_ind, F = ten_ind)) %>% 
  unite(col = "cross", sep = " x ", remove = F) %>% 
  spread(key = F, value = cross) %>% 
  dplyr::rename(Parents = M)

mating_mat[,-1][upper.tri(mating_mat[,-1], diag = TRUE)] = NA
mating_mat[c(9, 8, 5, 5), c(2, 5, 8, 4)] = NA
# mating_mat[sample(1:10, 2, replace = TRUE), sample(1:10, 3, replace = TRUE)]
  
mating_mat %>%
  knitr::kable(caption = "Partial diallel mating scheme using 10 parents") %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", font_size = 10) %>% 
  kableExtra::column_spec(1, bold = TRUE) %>% 
  kableExtra::column_spec(column = 1:11, width = "7em")
```


4. Connected diallels: Two groups (for example, $1-6$ and $7-12$) of individuals are used to form two diallels (generally partial or other diallel scheme in each group) but they are connected by crossing $4 \times 9$, $7 \times 1$, $9 \times 3$ and $10 \times 2$. In the example below, the second diallel also includes some selfs $(S)$ and reciprocals $(R)$ (Table connected-diallel).

With the convenience of using individuals as both male and female parents, diallel mating designs are popular for plant breeding studies. Certain diallel designs allow for estimation of reciprocal cross effects. Diallels cannot be used in dioecious species (female and male flowers occur in different plants). However, factorial designs can be used in dioecious species to estimate dominance genetic variance.

If there are no connections between groups of parents, the design is a diallel in sets. Diallel mating designs provide good evaluation of parents and full-sib families. They also provide estimates of both additive and dominance genetic effects, and genetic gains due to additive and dominance genetic effects if we assume the sample of parents used is sufficient to represent the reference population [@baker1978issues; @holland2003estimating]. One disadvantage of diallels is that the breeding and progeny evaluations can be costly due to large number of crosses required. As evident, For a full diallel with 6 parents, 36 crosses are required; with 12 parents the number of crosses required is `r 12^2`. On the other hand there is the looming question, if the sample of 6 or 12 individuals from a population provides useful estimate of the reference population genetic variances [@baker1978issues]. @white2007forest and @hallauer1988quantitative have described several other forms of diallel mating designs in detail.


<!-- ## Use of sparse matrices in mating designs -->

```{r sparse-matrices, echo=FALSE, include=FALSE}
# One may first learn the concept and use of sparse matrix
flev <- c(1,3:8)
Matrix::sparseMatrix(seq_along(flev), flev, x = 1)

```


## Specific combining ability (SCA) and Genetic Values

Individual phenotype ($y_i$) is:

$$y_i = \mu + A_i + D_i + e_i$$

SCA effects are derived from inter-allelic/intra-loci interactions. The mean genotypic value of offspring from a particular cross may deviate from value expected considering the population mean and the sum of the parental GCA effects. This deviation is the specific combining ability for that cross.

We can define the mean genotypic value ($G_AB$) for the full-sib family produced by crossing parents A and B as the sum of the overall mean $\mu$, the GCAs of the two parents and the SCA value:

$$G_AB = \mu + GCA_A + GCA_B + SCA_AB$$

A classical method to estimate dominance genetic variance (D) is to estimate the variance associated with SCA effects of many crosses. The expected value of the observed SCA variance component is 1/4 of the dominance genetic variance in the reference population.

## References
