---
title: Grade X result
author: Deependra Dhakal
date: '2018-10-19'
link-citations: yes
math: yes
slug: grade-x-result
categories:
  - tidyverse
  - R
tags:
  - tidyverse
  - R
header:
  caption: ''
  image: ''
---

<style type="text/css">

h1.title {
  font-size: 50px;
  color: DarkRed;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
                      message = FALSE, cache = FALSE, 
                      fig.align = "center")
require(tidyverse)
options(knitr.kable.NA = "", digits = 3)
```

---

<center>

## Total Marks and Final Grades {-}

```{r import}
result_9 <- read_csv("./data/score_sheet.csv")

org_colname <- result_9 %>% 
  colnames() %>% 
  stringr::str_subset("X.*")

paste_with1 <- result_9[2,] %>% 
  t() %>% 
  map_chr(c(1)) %>% 
  unique() %>% 
  magrittr::extract(-1)

paste_with2 <- result_9[1, ] %>% 
  map_chr(1) %>% 
  set_names(nm = NULL) %>% 
  unique() %>% 
  magrittr::extract(-1)

final_name44 <- expand.grid(paste_with1, paste_with2) %>% 
  select(rev(everything())) %>% 
  unite(col = paste12) %>% 
  pull(paste12)

# contains scores in upto X34
result_9 <- result_9 %>% 
  rename_at(vars(`Subject`:`X34`), function(x)x <- final_name44)

# scoresheet columns
# English_Theory:`O. Maths_Aggregate Grades`

# split the character coded marks of each student to a list
result_9 <- result_9 %>% 
  slice(-c(1, 2)) %>% 
  mutate_at(vars(contains("Theory"), contains("Practical")), .funs = function(x)str_split(x, pattern = "\\+", simplify = FALSE))

# change the list to numeric type for algebraic operation
result_9 <- result_9 %>% 
  mutate_at(vars(contains("Theory"), contains("Practical")), 
            function(x)map(x, ~ as.numeric(unlist(.x))))

# give a custom function to aggregate each marks of students
sumthree <- function(x){x[1]*.1 + x[2]*.3+x[3]*.6}
result_9 <- result_9 %>% 
  mutate_at(vars(contains("Theory"), contains("Practical")), 
            function(x)map_dbl(x, sumthree))

# mutate manually
result_9 <- result_9 %>% 
  mutate(`English_Total Marks` = English_Theory+English_Practical, 
         `Nepali_Total Marks` = Nepali_Theory+Nepali_Practical,
         `C. Maths_Total Marks` = `C. Maths_Theory`, # no practical
         `Science_Total Marks` = Science_Theory+Science_Practical,
         `Extension_Total Marks` = Extension_Theory+Extension_Practical,
         `Soil_Total Marks` = Soil_Theory+Soil_Practical,
         `Plant protection_Total Marks` = `Plant protection_Theory`+`Plant protection_Practical`,
         `Fruit_Total Marks` = Fruit_Theory+Fruit_Practical,
         `Agronomy_Total Marks` = Agronomy_Theory+Agronomy_Practical,
         `Computer_Total Marks` = Computer_Theory+Computer_Practical,
         `O. Maths_Total Marks` = `O. Maths_Theory` # no practical
  )

# colnames(result_9)[-c(1, 2, 47)] %>%
#   stringr::str_replace_all(c("_Total Marks|_Practical|_Theory|_Aggregate Grades"), "") %>% unique()

result_9 <- result_9 %>% 
  mutate_at(vars(contains("Total")), 
            funs(Agg_grade = case_when(
    . >= 90 & . < 100 ~ "A+",
    . >= 80 & . < 90 ~ "A",
    . >= 70 & . < 80 ~ "B+",
    . >= 60 & . < 70 ~ "B",
    . >= 50 & . < 60 ~ "C+",
    . >= 40 & . < 50 ~ "C",
    . >= 20 & . < 40 ~ "D",
    . >= 1 & . < 20 ~ "E",
    . >= 0 & . < 1 ~ "N",
    TRUE ~ as.character(`.`)
  )))

color_composer <- function(x){
  kableExtra::cell_spec(x, 
                        "html", 
                        color = case_when(is.na(x) ~ "white",
                                          x >= 40 ~ "blue",
                                          x < 40 ~ "red"
                        ))
}

bg_composer <- function(x)kableExtra::cell_spec(x, 
                                                "html", 
                                                color = case_when(
                                                  is.na(x) ~ "white",
                                                  TRUE ~ as.character(x)
                                                  ),
                                                background = case_when(
                                                  is.na(x) ~ "", # white
                                                  x %in% c("D", "E", "N") ~ "burlywood",
                                                  !x %in% c("D", "E", "N") & !is.na(x) ~ "aliceblue"
                                                ))

```


```{r printable-kable, results="asis"}

result_9 %>% 
  select_at(vars(1, contains("Total"), contains("Agg"))) %>% 
  rename_all(.funs = function(x)stringr::str_replace_all(x, "_", " ")) %>% 
  mutate_at(vars(ends_with("Total Marks")), color_composer) %>% 
  mutate_at(vars(contains("Agg")), bg_composer) %>% 
  knitr::kable(format = "html", 
               booktabs = TRUE, 
               align = "c", escape = FALSE,
               longtable = TRUE) %>% 
  kableExtra::kable_styling(position = "center", font_size = 14, 
                            bootstrap_options = c("striped", "hover")) %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::column_spec(1, width = "1.1cm", bold = TRUE) %>% 
  # kableExtra::column_spec(2:13, width = "1.3cm") %>% 
  kableExtra::landscape()

# # latex format
# result_9 %>% 
#   select_at(vars(1, 2, contains("Aggregate"))) %>% 
#   rename_all(.funs = function(x)stringr::str_replace_all(x, "_", " ")) %>% 
#   knitr::kable(format = "latex", 
#                booktabs = TRUE, 
#                align = "c", 
#                longtable = TRUE) %>% 
#   kableExtra::kable_styling(position = "center", font_size = 10, latex_options = c("striped", "Hold_position")) %>% 
#   kableExtra::row_spec(0, bold = TRUE) %>% 
#   kableExtra::column_spec(1, width = "1.1cm") %>% 
#   kableExtra::column_spec(2, width = "3.0cm") %>% 
#   kableExtra::column_spec(3:13, width = "1.3cm") %>% 
#   kableExtra::landscape()
```

**NA** indicates the absence of student

---

Formula used for calculating Aggregate GPA:

\begin{aligned}
Aggregate\ GPA = \frac{1^{st}\times 10 + 2^{nd}\times 30 + 3^{rd}\times 60}{100}
\end{aligned}

---

## Subjectwise marks distribution of all students {-}

```{r visualization-pass, fig.asp=0.8, fig.width=6, warning=FALSE, message=FALSE, fig.cap="Subjectwise percentage distribution of grade 9 Final examination"}
# # tidy gather, extract and spread
subject_wise <- result_9 %>%
  gather(key = Subject, value = Score,
         -`Roll Number`) %>%
  extract(col = Subject, into = c("Subject", "Th_Pr"), "^(.*)_(Theory|Practical)") %>%
  mutate(Score = as.numeric(Score)) %>%
  group_by(`Roll Number`, Subject, Th_Pr) %>%
  summarise_at(vars(Score), function(x)mean(x, na.rm = TRUE)) %>%
  spread(key = Subject, value = Score)

subject_wise %>% 
  filter(!is.na(Th_Pr)) %>% 
  select(-`<NA>`) %>% 
  gather(key = "Subject", value = "Scores", -Th_Pr, -`Roll Number`) %>% 
  group_by(Th_Pr, Subject) %>% 
  summarise_at(vars(Scores), function(x)mean(x, na.rm = TRUE)) %>% 
  ggplot(aes(as.factor(Subject), Scores)) +
  geom_col(position = "stack", aes(fill = Th_Pr)) +
  coord_flip() +
  ylab("Percentage Score") +
  xlab("Subject") +
  labs(fill = "Course type\n")
```

</center>

