---
title: Simulating genetic drift
author: Deependra Dhakal
date: '2020-11-08'
slug: simulating-genetic-drift
categories:
  - R
tags:
  - population genetics
  - R
subtitle: ''
summary: ''
authors: []
lastmod: '2020-11-08T07:55:52+05:45'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=4, 
                      echo=TRUE, warning=FALSE, 
                      message=FALSE, cache = TRUE)

require(tidyverse)
require(janitor)
require(viridis)
require(viridisLite)

theme_set(theme_bw())
```

Genetic drift is the result of bernouli process on survival of individuals (given some probability for each of them) of a population over a number of independent trials (Generation).

Apparently there are two techniques of seeing such process -- one individual level, other the population level. Both solutions are illustrated below. Let us suppose population of N individuals remains fixed from generation to generation, likewise, Fitness probability of "A" allele ($p(A)$) and "a" allele ($p(a)$) both starts off equal. Now we can generate incremental population survival probability for each individual for given population size:

```{r}

N <- 100
pA <- vector()
pA[1] <- 0.5
i <- 1

while ((pA[i] < 1) & (pA[i] > 0)) {
  nA <- 0

  for (j in 1:N){
    random <- runif(1)
    if(random < pA[i]){nA <- nA + 1}
  }
  pA[i + 1] <- nA/N
  i <- i + 1
}
```

Alternatively `rbinom` function generates the same but with probabilistic draw from entire population.

```{r}
drift_generate <- function(N = 100){
  N <- N
  pA <- vector()
  pA[1] <- 0.5
  i <- 1
  
  while ((pA[i] < 1) & (pA[i] > 0)) {
    nA <- rbinom(n = 1, size = N, prob = pA[i])
    pA[i + 1] <- nA/N
    i <- i + 1
  }
  return(tidyr::tibble(i = 1:i, pA = pA))
}

drift_tibble <- purrr::map_dfr(c(pop1 = 1, pop2 = 2, 
                                 pop3 = 3, pop4 = 4, 
                                 pop5 = 5, pop6 = 6), 
                               ~drift_generate(N = 100), .id = "population")

drift_gg <- ggplot(aes(x = i, y = pA), data = drift_tibble) +
  # geom_point(aes(color = population)) +
  geom_path(aes(color = population), size = 1.0) +
  theme_light() +
  scale_color_viridis_d() +
  labs(x = "Generation")

drift_gg
```

