---
title: 'Array operation: Outer product'
author: Deependra Dhakal
date: '2020-07-31'
slug: array-operation-outer-product
categories:
  - R
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2020-07-31T22:06:20+05:45'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Incidentally, I ran into outer product function `outer` today. It is extremely powerful function, in that it computes all combinations of product of two objects. One simplest and obvious demonstration is the multiplication table of numbers. :smile:

We can show multiplication table of numbers 1 through 12 each multiplied 1 through 10.

```{r}

outer(1:10, 1:12, "*")
```

Another immediate use can be seen in design and combinatorics. For instance,

```{r}
# outer product sum and subtraction function
outer(1:5, 1:10, "+")
outer(1:5, 1:10, "-")
```

Interestingly, outer subtraction of same two numeric vector gives a symmetric matrix with negative elements in upper with respect to lower.

```{r}
outer(1:5, 1:5, "-")
table(outer(1:5, 1:5, "-"))
```

In genetics, outer product is useful finding genotype combinations from multiple loci. For two loci, combination of genotypes is obtained as:

```{r}
outer(c("AA", "Aa", "aa"), c("BB", "Bb", "bb"), paste0)
```

Likewise for gamete formation, we can see that when there are 5 alleles in a diploid organism, possible genotype combinations are:

```{r}
outer(LETTERS[1:5], LETTERS[1:5], paste0)
```

