---
title: Tidyverse and tidbits
author: Deependra Dhakal
date: '2018-10-18'
slug: tidyverse-and-tidbits
categories:
  - R
  - tidyverse
tags:
  - R
  - tidyverse
output:
  blogdown::html_page:
    toc: true  
header:
  caption: ''
  image: ''
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("rlang")
library("tidyverse")
library(gapminder)
library(gridExtra)
```

# Ideas surrounding tidy evaluation

1. R code is a tree

    Every expression in R can be broken down to a form represented by a tree. For instance, on top of the tree there is "a function call" followed by it's branches: first child = function name itself, other children = function arguments. Complex calls have multiple levels of branching.

2. The code tree could be captured by quoting

    expr() quotes **your**(function developer) expression
    
    f1 <- function(z) expr(z)
    
    f1(a + b)
    
    enexpr() quotes **user's** expression
    
    f2 <- function(z) enexpr(z)
    
    f2(x + y)

3. Quoting and unquoting of expression enables building up of complex code trees
    
    Unquoting is achieved by bang bang operator (!!)

For example, let's create a function that makes scatter plot when fed a dataframe, x-variable and y-variable:
    
```{r scatter-maker}
scatter_maker <- function(df, xvar, yvar) {
  xvar <- enexpr(xvar)
  yvar <- enexpr(yvar)
  
  ggplot(df, aes(!!xvar, !!yvar)) +
    geom_point()
}
```

The function is executed in the demo below:

```{r scatter-demo, fig.cap="Scatterplot of displacement versus miles per gallon of mtcars dataset"}
scatter_maker(mtcars, disp, mpg) + 
  theme_bw()
```

4. Quosures capture both expression and environment

# quote() and eval()

1. Quoting an expression is a bit like quoting a string

```{r quote-example}
hm_quoted <- quote(height / mass)
hm_quoted
```

2. But it's R code that can be evaluated and with a data mask it works!

```{r quote-evaluation}
# eval(hm_quoted) # doesn't work because lacks data mask
head(eval(hm_quoted, starwars)) # works
```

# Why quote at-all

1. Data masking
2. Allows you to refer to data objects directly

```{r quoting-benefits}
# because transmute() is a quoting function,
transmute(starwars, height / mass)

# # unlike below where data objects could not be directly refered to
# height / mass
# starwars$height / starwars$mass
```

# Resolved messy data with tidy evaluation

A messy dataset exists despite all the sufferings. No wonder, it is resolvable, though. 

The identification is done through keywords like "gen1", "gen2", ... The rowtype and the height columns occur in precedence to the identifier key. Our function will break apart the respective entries of rowtype and height into separate columns.

```{r messy-dat}
dat <- tibble(mess = c("row:Three-row", "ht:Short height", "gen1", 
                       "row:Two-row", "ht:Tall height", "gen2",
                       "row:Three-row", "ht:Short height", "gen4",
                       "ht:Tall height", "row:Three-row", "gen3",  
                       "row:Three-row"))

dat
```

Let's define new verb `untangle2()`.

```{r untangler}
untangle2 <- function(df, regex, orig, new) {
  orig <- enexpr(orig)
  new <- sym(quo_name(enexpr(new)))
  # orig <- enquo(orig) # this too works!
  # new <- sym(quo_name(enquo(new))) # this too works!
  
  df %>% 
    mutate(
      !!new := if_else(grepl(regex, !! orig), !! orig, NA_character_)
    ) %>% 
    fill(!! new) %>% 
    filter(! grepl(regex, !! orig))
}

dat %>% 
  untangle2("^row", mess, rowtype) %>% 
  untangle2("^ht", mess, height)
```


# List behavior

When supplied a list to a dataframe as argument, list elements will be attempted to be supplied as individual column vectors. As long as the length of list element is same as the number of rows in dataframe, columns will be appended without an error.

However, if the complete list object is to be added in as a column vector, it's identity needs to be preserved. Function `I()` preserves the identity by wrapping the object around "AsIs" class. For example, let's create a dataframe with a list object containing it's elements.

```{r list-equal-length}
# all list elements of equal length as `nrow`.
df1 <- data.frame(x = 1:3, y = list(3:5, 4:6, 5:7))
df

# this results in error due to an attempt to bind list elements of unequal length as columns
# df <- data.frame(x = 1:3, y = list(3:9, 1:3, 1:3))
```

When dealing with unequal length lists, try wrapping the list around in `I()` function, as shown below.

```{r list-unequal-length}
df <- data.frame(x = 1:3, y = I(list(3:9, 1:3, 1:3)))
df

# Distinction can be made based on the classes
purrr::compose(class)(list(3:9, 1:3, 1:3))
purrr::compose(class, I)(list(3:9, 1:3, 1:3)) # compose applies multiple functions right to left
```


# Using pmap to perform list operations inside a dataframe

```{r list-col-df-purrr}
params <- expand.grid(param_a = c(2, 4, 6), 
                      param_b = c(3, 6, 9), 
                      param_c = c(50, 100), 
                      param_d = c(1, 0))

params %>%
  knitr::kable(caption = "Fictious data set to be used as template", 
               format = "html") %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", font_size = 10)

# using map function to obtain a list
df.preprocessed <- as_tibble(params) %>% 
  mutate(test_var = purrr::map(param_a, function(x){rep(5, x)}))

# using pmap for more flexible operation and unnesting to get expanded list
df.preprocessed1 <- dplyr::as_tibble(params) %>%
  mutate(test_var = purrr::pmap(., 
                                function(param_a, param_b, ...){
                                  rep(5, param_a) * param_b
                                })) %>% 
  tidyr::unnest()

# rowwise operation will also enable the same
my_fun <- function(param_a, param_b){
  rep(5, param_a) * param_b
}

dplyr::as.tbl(params) %>%
  rowwise() %>% 
  dplyr::mutate(test_var = list(my_fun(param_a, param_b))) %>% 
  tidyr::unnest()

# of all pmap solution could be the most flexible option
dplyr::as_tibble(params) %>%
  dplyr::mutate(test_var = purrr::pmap(list(x = param_a
                                            ,y = param_b
                                            ,z = param_c
                                            ,u = param_d),
                                       ~ rep(5,..1)*..2)
  )
```


# Apply a function to certain columns only, by rows

```{r function-by-row}
mtcars %>% 
  select(am, gear, carb) %>% 
  purrrlyr::by_row(sum, .collate = "cols", .to = "sum_am_gear_carb")
```

# Do any arbitrary operation

```{r do-operation-on-groups}
mtcars %>% 
  group_by(cyl) %>% 
  dplyr::do(broom::tidy(lm(mpg ~ drat + wt, data = .)))
```

The do call is combined with group_by to get any summary of current dataframe

```{r do-operation-on-groups2}
iris %>%
  group_by(Species) %>%
  do({
    mod <- lm(Sepal.Length ~ Sepal.Width, data = .)
    pred <- predict(mod, newdata = .["Sepal.Width"])
    data.frame(., pred)
  })

# # same thing could be done through explicit function call with do
# iris %>%
#   group_by(Species) %>%
#   do((function(df) {
#     mod <- lm(Sepal.Length ~ Sepal.Width, data = df)
#     pred <- predict(mod,newdata = df["Sepal.Width"])
#     data.frame(df,pred)
#   })(.))

```


# Using dplyr functions inside your own functions

Create column selection/extraction function, for example.

```{r extending-dplyr-verbs}
extract_vars <- function(data, some_string){
  data %>% 
    select(contains({{ some_string }})) -> data
  
  return(data)
}

extract_vars(mtcars, "drat")
```


# Effiecient ways to operate on list

```{r demo-list1}
x <- list(
  list(name = "sue", number = 1, veg = c("onion", "carrot")),
  list(name = "doug", number = 2, veg = c("potato", "beet"))
)
```

Base `rbind()` returns a matrix of lists. However, a complete list object `x` is packed in a single row, with two list columns for each next level list.

```{r base-rbind}
rbind(x)
```

# Binds each element as row but third element is list

Throwing a `do.call()` call leads to desired results.

```{r do-calls-on-list}
do.call(rbind, x)
do.call(rbind, x) %>% class()
```


# Produces rowbinding for dataframe but doesn't work with list elements

```{r dfr-limitation}
map_dfr(x, ~ .x[c("name", "number")])
```

# Tibble and map data to list column in isolation

```{r tibble-specification}
tibble(
  name = map_chr(x, "name"), 
  number = map_dbl(x, "number"), 
  veg = map(x, "veg")
)
```


# Modeling, exploration, summary extraction and visualization examples using dplyr verbs.

```{r modeling-and-visualization}
gap <- gapminder %>%
  filter(continent == "Asia") %>%
  mutate(yr1952 = year - 1952)

ggplot(gap, aes(x = lifeExp, y = country)) +
  geom_point()

gap <- gap %>%
  mutate(country = fct_reorder2(country, .x = year, .y = lifeExp))

ggplot(gap, aes(x = lifeExp, y = country)) +
  geom_point()

gap_nested <- gap %>%
  group_by(country) %>%
  nest()

gap_nested$data[[3]]

gap_fitted <- gap_nested %>%
  mutate(fit = map(data, ~ lm(lifeExp ~ yr1952, data = .x)))
gap_fitted

gap_fitted <- gap_fitted %>%
  mutate(
    intercept = map_dbl(fit, ~ coef(.x)[["(Intercept)"]]),
    slope = map_dbl(fit, ~ coef(.x)[["yr1952"]]),
    r_squared = map_dbl(fit, ~ summary(.x)[["r.squared"]])
  )

gap_fitted %>% 
  top_n(n = 5, wt = r_squared) # whoa! Nepal is 4th

ggplot(gap_fitted, aes(x = intercept, y = country)) +
  geom_point()

ggplot(gap_fitted, aes(x = slope, y = country)) +
  geom_point()

```

To get summary statistics implementing certain custom functions on list items. The case example below shows application of some high level functions available in tidyverse.

```{r modeling-and-summary-extraction}
set.seed(1)

# construct a dataframe
d <- tibble(n = c(5, 1, 3))
e <- d %>% group_by(n) %>% 
  do(tibble(y = rnorm(.$n), dat = list(data.frame(a = 1)))) 

# this method is tranmutating type (drops original)
e %>% rowwise() %>% do(tibble(sum = .$y + .$n))

# this only adds the mutated summary (keeps original)
mutate(e, sum = purrr::map2_dbl(y, n, `+`))

# # .f should be able to work with a list or a data frame. As it
# happens, sum() handles data frame so the following works:
mtcars %>% purrrlyr::by_row(sum)

# # Other functions such as mean() may need to be adjusted with one
# # of the lift_xy() helpers:
mtcars %>% purrrlyr::by_row(purrr::lift_vl(mean))

# To run a function with invoke_rows(), make sure it is variadic (that
# it accepts dots) or that .f's signature is compatible with the
# column names
mtcars %>% purrrlyr::invoke_rows(.f = sum)
mtcars %>% purrrlyr::invoke_rows(.f = purrr::lift_vd(mean))

# # invoke_rows() with cols collation is equivalent to plyr::mdply()
p <- expand.grid(mean = 1:5, sd = seq(0, 1, length = 10))
p %>% purrrlyr::invoke_rows(.f = rnorm, n = 5, .collate = "cols")

# p %>% plyr::mdply(rnorm, n = 5) %>% dplyr::tbl_df()

# # To integrate the result as part of the data frame, use rows or
# # cols collation:
mtcars[1:2] %>% purrrlyr::by_row(function(x) 1:5)
mtcars[1:2] %>% purrrlyr::by_row(function(x) 1:5, .collate = "rows")
mtcars[1:2] %>% purrrlyr::by_row(function(x) 1:5, .collate = "cols")

# # getting custom summary functions with purrr and dplyr verbs
mtcars %>% 
  group_by(cyl) %>% 
  tidyr::nest() %>% 
  mutate(model = purrr::map(data, ~lm(qsec ~ hp, .)),
         coef  = purrr::map_dbl(model, ~coef(.)[2])) %>% 
  tidyr::unnest(data)

# # run a regression after group_by operation
# # and save results to dataframe
group_by(mtcars, gear) %>% do(res = residuals(lm(mpg~carb, .)))

# # yet better way to capture the results -- residuals.
# # then it can be combined with original data. 
mtcars %>%
  group_by(cyl) %>%
  do(model = lm(mpg ~ wt, data=.)) %>%
  do((function(reg_mod) {
    data.frame(reg_res = residuals(reg_mod$model))
  })(.))

# # Custom data modeling
# # using dplyr do() calls
fertility <- read_csv(here::here("content", "blog", "data", "fertility.csv"))

# # binding predictions to original dataframe
fertility %>%
  group_by(Year) %>%
  do(data.frame(.,pred = predict(mgcv::gam(fertility ~ s(AGE), data=.))))

# # same but different approach
do.call(rbind,
        lapply(split(fertility, fertility$Year), function(df) {
          df$pred <- predict(mgcv::gam(fertility ~ s(AGE), data=df))
          df
        }))

# # same using data.table library
fertility %>%
  data.table::data.table() %>%
  group_by(Year) %>%
  mutate(pred = predict(mgcv::gam(fertility ~ s(AGE))))

# # this fails at last line
# fertility %>%
#   group_by(Year) %>%
#   dplyr::do(model = mgcv::gam(fertility ~ s(AGE), data = .)) %>%
#   left_join(fertility, .) %>%
#   mutate(smoothed = predict(model, newdata = AGE))

# # multiple model fitting and parameter extraction
fitmodel <- function(d) lm(mpg ~ disp, data = mtcars)
by_cyl <- mtcars %>% 
  group_by(cyl) %>% 
  tidyr::nest() %>%
  mutate(mod = purrr::map(data, fitmodel), 
         pars = purrr::map(mod, broom::tidy), 
         pred = purrr::map(mod, broom::augment))

pars <- by_cyl %>% tidyr::unnest(pars)
pars
prediction <- by_cyl %>% tidyr::unnest(pred)
prediction
```

# Workarounds on unequal length lists

```{r dealing-unequal-length}
hareramshyam <- list(harry = c(1, 2, 3, 4, 5), ram = c(2, 3, 4, 6, 8), shyam = c(3, 4, 5, 3, 2, 4, 5, 6, 3))

## binding rows for two list items of equal lengths is possible
bind_rows(hareramshyam[1:2])

# bind_rows(hareramshyam) ## this fails

## so to complete the list by filling with nas to match the 
## length of each element in list, use
mymax <- hareramshyam %>% 
  lapply(length) %>% 
  unlist %>% 
  max

hareramshyam %>% 
  lapply(function(x) c(x, rep(NA_integer_, mymax-length(x)))) %>% 
  bind_rows()

## here is a two liner that could be cause the same
lapply(hareramshyam, `length<-`, 
       max(lengths(hareramshyam))) %>% 
  bind_rows()

# or use plyr's binder to avoid recycling
plyr::rbind.fill(map(hareramshyam, ~ as.data.frame(t(.x),stringsAsFactors=FALSE)))

## to fill, for example above, dataframe of NAs with other
## elements of the dataframe
lapply(hareramshyam, `length<-`, 
       max(lengths(hareramshyam))) %>% 
  bind_rows() %>% 
  mutate(harry = dplyr::coalesce(harry, shyam), 
         ram = dplyr::coalesce(ram, shyam))
```

