---
title: Missing negative from the normal
author: Deependra Dhakal
date: '2020-11-07'
slug: missing-negative-from-the-normal
categories:
  - R
tags:
  - fun
  - R
subtitle: ''
summary: ''
authors: []
lastmod: '2020-11-07T19:37:02+05:45'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
math: yes
---

# A normal function isn't so normal

The normal density function is:

$$
\large
f(x) = \frac{1}{\sqrt{2 \pi} \sigma} \exp^{-\frac{(x - \mu)^2}{(2 \sigma^2)}}
$$

It doesn't make sense to calculate the probability for a single value in a continuous probability function, it is by definition zero, but you can calculate relative likelihoods (heights). `dnorm` simply gives the value of the function for a given x, not the area under the curve for that x (which is basically nothing for a single value). To find the density (height) for a single x value on the normal distribution, use `dnorm()` in the following way (here each x value is treated as separate and vectorized over),

https://stackoverflow.com/questions/57345406/estimating-probability-density-in-a-range-between-two-x-values-on-simulated-data

```{r}
example_data <- c(0.2, 1.2, 1.2, 1.4, 1.5, 1.3, 1.2, 1.3, 1.6, 1.9, 2.3)

dnorm(example_data, mean = 0, sd = 1)
normal_density <- function(x, mu=0, sd=1){1/(sqrt(2 * pi)*sd) * exp(-((x - mu)^2/(2 * sd^2)))}
normal_density(example_data, mu = 0, sd = 1)
plot.function(normal_density, from = example_data)
```

It is also possible to obtain approximate probability density function using `density` and `approxfun` 

```{r}
estimated_dfun <- approxfun(density(example_data), rule=2)
```

From the estimated density function point probability density estimates could be made as:

```{r}
estimated_dfun(1.45) # use any number to find the height ?

# the area under the curve:
area_under <- function(lower, upper) {
    integrate(estimated_dfun, lower=lower, upper=upper) }
```


ecdf() is the way to go for a non-parametric estimate, but if you expect a normal distribution you can also do a parametric estimate.

```{r}
dist_mean <- 10
dist_sd <- 0.2
a <- 9.7
b <- 10.2

set.seed(123)
r <- rnorm(1e4, dist_mean, dist_sd)

# population
pnorm(b, dist_mean, dist_sd) - pnorm(a, dist_mean, dist_sd)
# [1] 0.7745375

# parametric estimate
pnorm(b, mean(r), sd(r)) - pnorm(a, mean(r), sd(r))
# [1] 0.7753985

# nonparametric estimate
ecdfun <- ecdf(r)
ecdfun(b) - ecdfun(a)
# [1] 0.7754
```


## Bootstrap tutorial

https://matthew-parker.rbind.io/post/2020-06-26-bootstrap-tutorial/

## Probability distribution

https://jasonbintz.rbind.io/courses/math333/notes/probability-distributions/

## Linear regression illustrated

https://vinaysingh.rbind.io/2020/03/09/linear-regression/

## Mixed model simulation

https://aosmith.rbind.io/2018/04/23/simulate-simulate-part-2/
